The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-07-11 paper.tex
Added further exposition on Sonority Angle

In the diagram in \ref{sonangle_picture}, imagine fixing C\textsubscript{1} and lowering the sonority of C\textsubscript{2}. Intuitively, this increases the {\sc Sonority Angle}. The dependence of {\sc Sonority Angle} on this distance can also be seen in the second term in \ref{sonangle_formula}.

>>>>

Let us verify that {\sc Sonority Angle} does indeed reflect the two generalisations we wish to make:
first, that the smaller the sonority distance between C\textsubscript{1} and C\textsubscript{2},
the smaller the sonority angle.
Imagine fixing C\textsubscript{1} as in \ref{sonangle_picture} and raising the sonority of C\textsubscript{2}. Intuitively, this decreases the {\sc Sonority Angle}, comparing the two below.

\ex. \begin{tikzpicture} [shorten >=1pt,scale=0.33]
                      \draw [<-] (-3,6.5) -- (-3, 0.5) ; % axis
                      \node at (-3, 7.0) {Sonority}; % axis label
					 % left diagram
                      \draw (5,1) -- (0,3) -- (5,6) ; % rising sonority
                      \node at (1.8, 3.2) {$\theta$}; % theta label for rising sonority
                      \node[left] at  (0,3) {C$_1$}; 
                      \node[right] at (5,1) {C$_2$};
                      \node[right] at (5,6) {V};
					  % right diagram
                      \draw (20,4) -- (15,3) -- (20,6) ; % falling sonority
                      \node at (17.5, 4.0) {$\theta$}; % theta label for falling sonority
                      \node[left] at (15,3) {C$_1$}; 
                      \node[right] at (20,4) {C$_2$};
                      \node[right] at (20,6) {V};
\end{tikzpicture}

The dependence of {\sc Sonority Angle} on this distance can also be seen in the second term in \ref{sonangle_formula}.

The second generalisation is that the more sonorous the C\textsubscript{1}, the more likely
epenthesis is to occur. This time, fix C\textsubscript{2} and lower the sonority of C\textsubscript{1}.

\ex. \begin{tikzpicture} [shorten >=1pt,scale=0.33]
                      \draw [<-] (-3,6.5) -- (-3, 0.5) ; % axis
                      \node at (-3, 7.0) {Sonority}; % axis label
					  % left diagram
                      \draw (5,1) -- (0,3) -- (5,6) ; % rising sonority
                      \node at (1.8, 3.2) {$\theta$}; % theta label for rising sonority
                      \node[left] at  (0,3) {C$_1$}; 
                      \node[right] at (5,1) {C$_2$};
                      \node[right] at (5,6) {V};
					  % right diagram
                      \draw (20,1) -- (15,1) -- (20,6) ; % falling sonority
                      \node at (17, 1.9) {$\theta$}; % theta label for falling sonority
                      \node[left] at (15,1) {C$_1$}; 
                      \node[right] at (20,1) {C$_2$};
                      \node[right] at (20,6) {V};
\end{tikzpicture}

While the difference is less clear visually, the second angle is larger. The first term in the formula confirms the relation between the sonority of C\textsubscript{1} in terms of its closeness to V, and {\sc Sonority Angle} as a whole.

Given a sonority scale where classes of consonants are mapped to a numerical sonority, we can now calculate
the {\sc Sonority Angle} for any cluster, which can be thought of as the faithfulness cost of epenthesising
between the two consonants.  Examples of the calculation are given below.

In this paper, I adopt (with, later, minor modifications) the following standard scale:

\ex. \label{standardsonorityscale} 
      \begin{tabular}{cccccc}
         T & F & N & R & G & V \\
         stop & fricative & nasal & liquid & glide & vowel \\
         1 & 2 & 3 & 4 & 5 & 6 \\
      \end{tabular}

The {\sc Sonority Angles} for NT, TT and TN are calculated as follows.

\ex. \a. \textsc{SonAngle}(NT) =  $arctan(6-3) - arctan(1-3)$ = 2.35
     \b. \textsc{SonAngle}(TT) =  $arctan(6-1) - arctan(1-1)$ = 1.37
     \c. \textsc{SonAngle}(TN) =  $arctan(6-1) - arctan(3-1)$ = 0.27

The larger the {\sc Sonority Angle}, the larger the faithfulness cost. We therefore expect it to be hardest to epenthesise into NT out of these three clusters, and easiest to epenthesise into TN.

We formalise the idea of the faithfulness cost by defining a family of \textsc{Ident} constraints that penalise outputs that incur faithfulness costs of greater than a certain $n$.

\ex. \textsc{Ident(SonAngle)}$<n$: Assign a violation mark if the consonants in two strings C\textsubscript{1}C\textsubscript{2} and C\textsubscript{1}VC\textsubscript{2} 
stand in correspondence, and the sonority angle between C\textsubscript{1}C\textsubscript{2} and C\textsubscript{1}V is greater than $n$. % TODO: change to it being a property of C1?

These faithfulness constraints have a universal ranking, with the least stringent the highest-ranked.

   \ex. ... \newline
            $\gg$ \textsc{Ident(SonAngle)}$<$1.5 \newline
            $\gg$ \textsc{Ident(SonAngle)}$<$1.0 \newline
            $\gg$ \textsc{Ident(SonAngle)}$<$0.5 \newline
            $\gg$ ...

Note that I do not claim that this is the only dimension of perceptual similarity relevant to epenthesis.  
As we will see later in the paper, other faithfulness constraints may prevent epenthesis, such as a restriction on epenthesising a vowel between two obstruents, which introduces an additional region of sonorance that had previously not existed \cite{flemming.2008}.
